---
title: "Demos from O'Reilly's Learning R (Richard Cotton)"
author: "Steven Troxler"
date: "December 1, 2015"
output: html_document
---

# Reading and writing data

Almost all R tutorials discuss `read.table`, `write.table`, `read.csv`, and `write.csv` for
delimited tables. Here we discuss some of the other forms of data IO covered by *Learning R*.

## Plain text

Use `readLines` to read plain-text files that can fit in memory; this is akin to `python`s
`with open(path, 'r') as f: f.readlines()`:
```{r}
setwd("~/lp/R/learningr/")
plain_text_path <- system.file("extdata",
                               "Shakespeare's The Tempest, from Project Gutenberg pg2235.txt",
                               package="learningr")
my_text <- readLines("a_plain_text_file.txt")
my_text
```

We can write using `writeLines`, which prints to stdout by default but writes
in "w" mode if you give it a path
```{r}
writeLines(my_text, "another_text_file.txt")
```

To write a single large string to a file, use `cat()`, which like `writeLines`
defaults to `stdout` but optionally accepts a path as its second argument.

## XML

When you read XML using the `XML` library, you get back documents with a
special type. If you add `useInternalNodes = FALSE` to the call, you'll be
able to use `str` and `head` to view the document. That said, the output is
extremely verbose, so we suppress the output here
```{r, message=FALSE}
xml_path <- system.file("extdata", "options.xml", package="learningr")
r_options <- XML::xmlParse(xml_path)
# can't use str(r_options) / head(r_options)
r_options <- XML::xmlParse(xml_path, useInternalNodes = FALSE)
str(r_options)
head(r_options)
```

There is also an `xmlTreeParse` function, which is like `xmlParse` with
`useInternalNodes = FALSE` by default.

You can work with XML using the XPath language, for example
```{r}
xpathSApply(r_options, "//variable[containes(@name, 'warn')]")
```
which searches the options for any variables whose name attribute
containes the string 'warn'.

If you need to write R data to an xml file (for example so that another
tool or language can read it), the `XML` package won't help, but
`RUniversal::makexml` will: you can apply it to a list and it will
output a string; you can then use `cat` to write it to a file.

## Json

There are two main packages for working with json, `rjson` and `RJSONIO`.
Since they are similar (they differ in how they handle NA / NaN), I'll just
demo `rjson`, which prefers to work with in-memory strings (`RJSONIO`, like
`XML`, likes to read directly from file).

TODO:
1. DEMO A SIMPLE TO-AND-FROM JSON WITH LISTS AND VECTORS, INCLUDING THE WAY
THAT IT HANDLES NAN AND NA.
2. DEMO DUMPING A DATA.FRAME AS JSON AND LOADING IT BACK UP DIRECTLY
3. DISCUSS THE IDEA OF RECORDS (MENTION THAT PANDAS HAS IT BUILT-IN), AND
DEMO DUMPING A DATA FRAME AS RECORDS (WHICH IS WHAT YOU'D WANT IN, SAY, D3)

## Yaml

You can read yaml from a file using `yaml::load_file(file_path)`; you can
similarly load yaml strings. You can dump an R object to yaml using
`yaml::as.yaml`.

## Excel files

## HDF5 archives

## SQL

# Plotting

I have already made examples of most of the plots from the plotting
chapter of *Learning R*, but I do want to go through the demonstration of
a bar chart with two independent variables.

## Modelling

TODO: decide whether this is even worth doing... *Learning R* isn't really
the book I would choose to use for learning modelling, on the other hand the
examples are decent.
