---
title: "Demos from O'Reilly's Learning R (Richard Cotton)"
author: "Steven Troxler"
date: "December 1, 2015"
output: html_document
---

# Reading and writing data

Almost all R tutorials discuss `read.table`, `write.table`, `read.csv`, and `write.csv` for
delimited tables. Here we discuss some of the other forms of data IO covered by *Learning R*.

## Plain text

Use `readLines` to read plain-text files that can fit in memory; this is akin to `python`s
`with open(path, 'r') as f: f.readlines()`:
```{r}
setwd("~/lp/R/learningr/")
plain_text_path <- system.file("extdata",
                               "Shakespeare's The Tempest, from Project Gutenberg pg2235.txt",
                               package="learningr")
my_text <- readLines("a_plain_text_file.txt")
my_text
```

We can write using `writeLines`, which prints to stdout by default but writes
in "w" mode if you give it a path
```{r}
writeLines(my_text, "another_text_file.txt")
```

To write a single large string to a file, use `cat()`, which like `writeLines`
defaults to `stdout` but optionally accepts a path as its second argument.

## XML

When you read XML using the `XML` library, you get back documents with a
special type. If you add `useInternalNodes = FALSE` to the call, you'll be
able to use `str` and `head` to view the document. That said, the output is
extremely verbose, so we suppress the output here
```{r, message=FALSE}
xml_path <- system.file("extdata", "options.xml", package="learningr")
r_options <- XML::xmlParse(xml_path)
# can't use str(r_options) / head(r_options)
r_options <- XML::xmlParse(xml_path, useInternalNodes = FALSE)
head(r_options)
```

There is also an `xmlTreeParse` function, which is like `xmlParse` with
`useInternalNodes = FALSE` by default.

You can work with XML using the XPath language, using the
`XML::xpathSApply` function. There's a lot of other xml functionality,
but most of it is hard to learn without a real problem to work on.

If you need to write R data to an xml file (for example so that another
tool or language can read it), the `XML` package won't help, but
`RUniversal::makexml` will: you can apply it to a list and it will
output a string; you can then use `cat` to write it to a file. (This,
the author points out, is how he made `options.xml` in the first place.)

## Json

There are two main packages for working with json, `rjson` and `RJSONIO`.
Since they are similar (they differ in how they handle NA / NaN), I'll just
demo `rjson`, which prefers to work with in-memory strings (`RJSONIO`, like
`XML`, likes to read directly from file).

To convert an R object to json, use `rjson::toJSON`:
```{r}
sample_data_frame <- data.frame(var1 = c(2, 3, 4), var1 = c("cat", "cat", "dog"))
json1 <- rjson::toJSON(
  list(x = c(1, 2, 4, NA), y = NaN, z = list(a = 5, b = "hello")))
json2 <- rjson::toJSON(sample_data_frame)
json1
```

Note how the data frame is interpreted as if it were a list of columns. This
is natural given that R treats data frames a bit like lists, but is not best
practice when working with other languages; we will come back to this.

To convert json back to R, use `rjson::fromJSON`:
```{r}
rjson::fromJSON(json1)
rjson::fromJSON(json2)
```

Note that to reconstruct our `NA` and `NaN` observations, we would need to
convert them explicitly. To reconstruct a data frame you dumped as json,
wrap the `fromJSON` call in an `as.data.frame`:
```{r}
as.data.frame(rjson::fromJSON(json2))
```

## Yaml

You can read yaml from a file using `yaml::load_file(file_path)`; you can
similarly load yaml strings. You can dump an R object to yaml using
`yaml::as.yaml`.

## Records

The way that R converts `data.frame`s to `list`s is natural given how R
works, but it is counterproductive for working with many other tools like
D3: these tools tend to expect tabular data to be formed as lists of maps.
In the `pydata` stack, the `DataFrame` class has methods `to_records` and
`from_records` to make this conversion easy.

Here, we define a `to_records` function for R. The `from_records` case is a
bit harder to write, and slightly less likely to be important so we omit it.

```{r}
to_records <- function(df) {
  records <- split(df, seq(nrow(df)))
  names(records) <- NULL
  records
}
```

## Excel files

## HDF5 archives

## SQL

# Plotting

I have already made examples of most of the plots from the plotting
chapter of *Learning R*, but I do want to go through the demonstration of
a bar chart with two independent variables.

# Modelling

TODO: decide whether this is even worth doing... *Learning R* isn't really
the book I would choose to use for learning modelling, on the other hand the
examples are decent.
