---
title: "Vignette Title"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, echo = FALSE}
# These options make it so that:
#   (a) output blocks are mixed with code, and prefixed with a comment
#   (b) figures are sized, by default, to use most of the html witdth
#       (you can reset fig.width and fig.height inside the {}'s at the
#       top of any code block, if you want a particular plot to be bigger
#       or smaller)
knitr::opts_chunk$set(
  collapse = TRUE, comment = '#>',
  fig.width = 7, fig.height = 6
)
```

# Getting a first look at the data

Let's explore data from the Kaggle Titanic competition
```{r}
library(slt.titanic)
dat <- load_clean_data()
names(dat)
train <- dat$train
```

First thing is to get a look at the raw data:
```{r}
str(train)
```

# Checking for duplicate records

Before we do anything else, let's put on our data science hats and
do a check for duplicate data (here I'm following along with David
Langer's youtube tutorial):
```{r}
library(dplyr)
combined <- dat$combined
name_cts <- combined %>%
            group_by(Name) %>%
            summarise(n_dups=n()) %>%
            arrange(desc(n_dups))
head(name_cts)
dup_names <- (name_cts %>% filter(n_dups > 1))$Name
dup_names
combined %>% filter(Name %in% dup_names)
```
From this, it looks like the duplicates are in fact different people who
happen to share the same name, rather than messy data.

# Taking a look at some of the categorical variables

Now that we've seen what the data looks like, let's glance
at a few of the univariate and bivariate distributions as tables:
```{r}
table(train$Survived)
table(train$Pclass, train$Survived)
table(train$Sex, train$Survived)
```

These tables show that there's definitely a connection between both survival
rate and both class and gender. Let's get a graphical look:
```{r}
library(ggplot2)
ggplot(train) + aes(x=Sex, fill=Survived) + 
                geom_bar() + facet_wrap(~ Pclass)
```

# Let's look at the titles in the names

A quick look at the names shows several different titles in use.
A quick look at some of the most common shows nothing very surprising;
many of them we could guess from the age and Sex, although the
distinctions between "Miss" and "Mrs" and also the use of "Don"/"Donna"
might give us some additional information.

```{r}
library(stringr)
title_table <- function(title) {
  table(str_detect(combined$Name, title), combined$Sex)
}
title_table("Mr\\.")
title_table("Miss\\.")
title_table("Mrs\\.")
title_table("Master")
title_table("Don\\.")
title_table("Donna\\.")
```

(Note that when using `stringr::str_detect`, we need to use "\\." to escape
the periods; the first backslash prevents R itself from interpreting the second
backslash, which tells the regexp matcher to treat the period as literal rather
than as "any character".)

We can also create a function (which we will probably add to our package
when we begin working on a full analysis) to extract the titles we saw from the
names:
```{r}
names_to_titles <- function(names) {
  titles = rep("Other", length(names))
  titles[str_detect(names, "Mr\\.")] = "Mr."
  titles[str_detect(names, "Mrs\\.")] = "Mrs."
  titles[str_detect(names, "Miss\\.")] = "Miss."
  titles[str_detect(names, "Master")] = "Master"
  titles[str_detect(names, "Don\\.")] = "Don."
  titles[str_detect(names, "Donna\\.")] = "Donna."
  as.factor(titles)
}
names_to_titles(combined$Name[1:10])
```

We can also investigate which titles we missed and improve
our function:
```{r}
combined$Name[is.na(names_to_titles(combined$Name))]
```